---
output:
  html_document: default
  word_document: default
---
White Wine Exploratory Data Analysis
========================================================

##Structure and introduction of the data set:
In this project I will explore and analyze the white wine dataset to see how the different input variables determine the quality of the white wine. First, we will start by looking at the distribution of each variable. Then, we will look at the correlation between the different variables. Finally, we will select three graphs to summarize our findings. Let's start by loading the data and have a brief look at the variables:


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(gridExtra)
library(plotrix)
library(GGally)
library(scales)
library(memisc)
library(corrplot)
```

Dimensions of the dataset:
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Loading the Dataset
white_wine<- read.csv('wineQualityWhites.csv')
white_wine$X<- NULL#removing the x var since it is not needed
dim(white_wine)
```
Input variables (based on physicochemical tests):<br/>
   1 - fixed acidity (tartaric acid - g / dm^3)<br/>
   2 - volatile acidity (acetic acid - g / dm^3)<br/>
   3 - citric acid (g / dm^3)<br/>
   4 - residual sugar (g / dm^3)<br/>
   5 - chlorides (sodium chloride - g / dm^3<br/>
   6 - free sulfur dioxide (mg / dm^3)<br/>
   7 - total sulfur dioxide (mg / dm^3)<br/>
   8 - density (g / cm^3)<br/>
   9 - pH<br/>
   10 - sulphates (potassium sulphate - g / dm3)<br/>
   11 - alcohol (% by volume)<br/>
Output variable (based on sensory data): <br/>
   12 - quality (score between 0 and 10)<br/>


Type of each variable:

```{r echo=FALSE, message=FALSE, warning=FALSE}
str(white_wine)
```
**summary of the variables:**

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(white_wine)
```

## Univariate Plots Section

In this section, and before we look at the relations between the variables and how they affect each other, We will look at each variable individually to have a better understanding of its structure and distribution shape. knowing the shape of the data will help us to have some expectations about how the graphs will be later on, and to check the outliers and remove them if needed.

let us start with the quality column:

```{r echo=FALSE, message=FALSE, warning=FALSE}
quality.proportions<- as.data.frame(table(white_wine$quality))#creating data frame from the quality scores and their count
quality.proportions<-quality.proportions[c(1,5,6,4,7,2,3),]# Rearraigning the quality scores to make them clearer 
piepercent<- round(100*quality.proportions$Freq/sum(quality.proportions$Freq), 1)# creating the percentage of each score
pie(quality.proportions$Freq,labels = piepercent, col = rainbow(7),
    main = "quality category proportions") 
legend("topright", quality.proportions$Var1, title = "quality rating",
       legend= quality.proportions$Var1,fill = rainbow(7))

```

> It seems like the quality rating 5 and 6 make the most of the qualities. Let us see the count of each category.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= quality), data= white_wine)+ geom_bar()
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
table(white_wine$quality)
```

> As we can see thier are very few high and low quality wines. So, we can create a new variable called **quality.buckets** with type **Factor**. Containing the following values.<br/>
  value    range<br/>
- low      (1-4]<br/>
- good     (4-6]<br/>
- high     (6-10]<br/>


```{r echo=FALSE, message=FALSE, warning=FALSE}
white_wine$quality.buckets<- cut(white_wine$quality,breaks = c(1,4,6,10),
                                 labels = c("low","good","high"))

ggplot(aes(x= quality.buckets), data= white_wine)+ geom_bar()
```


####let us have a look at input variables now:

```{r echo=FALSE, message=FALSE, warning=FALSE}

grid.arrange(ggplot(aes(x="alcohol", y= alcohol), data = white_wine)+geom_boxplot(),
qplot(x= alcohol, data= white_wine,bins = 50),ncol=2)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(white_wine$alcohol)
```

>If we look at the graph and the values of the median and the mean, we can see that alcohol distribution is slightly right-skewed.


```{r echo=FALSE, message=FALSE, warning=FALSE}
grid.arrange(ggplot(aes(x="sulphates", y= sulphates), data = white_wine)+geom_boxplot(),
qplot(x= sulphates, data= white_wine),ncol=2)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(white_wine$sulphates)
```

> The sulphates distribution is slightly right-skewed, I think if we removed the outliers, we will get a symmetric  distribution.

```{r echo=FALSE, message=FALSE, warning=FALSE}
grid.arrange(ggplot(aes(x="residual.sugar", y= residual.sugar), data = white_wine)+geom_boxplot(),
qplot(x= residual.sugar, data= white_wine),ncol=2)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(white_wine$residual.sugar)
```

>The distribution of the Residual Sugar is highly right-skewed, with an extreme outlier at 65.8 (g / dm^3). 

```{r echo=FALSE, message=FALSE, warning=FALSE}
grid.arrange(ggplot(aes(x="fixed.acidity", y= fixed.acidity), data = white_wine)+geom_boxplot(),
qplot(x= fixed.acidity, data= white_wine),ncol=2)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(white_wine$fixed.acidity)
```

> Again the graph is slightly right-skewed because of the outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE}
grid.arrange(ggplot(aes(x="volatile.acidity", y= volatile.acidity), data = white_wine)+geom_boxplot(),
qplot(x= volatile.acidity, data= white_wine),ncol=2)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(white_wine$volatile.acidity)
```

> The Volatile Acidity distribution is right-skewed with a lot of outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE}
grid.arrange(ggplot(aes(x="citric.acid", y= citric.acid), data = white_wine)+geom_boxplot(),
qplot(x= citric.acid, data= white_wine),ncol=2)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(white_wine$citric.acid)
```

> The distribution of the Citric Acid is slightly right-skewed. Also, the distribution have some extreme outliers and a peak at 0.26.

```{r echo=FALSE, message=FALSE, warning=FALSE}
grid.arrange(ggplot(aes(x="chlorides", y= chlorides), data = white_wine)+geom_boxplot(),
qplot(x= chlorides, data= white_wine,bins = 40),ncol=2)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(white_wine$chlorides)
```

>The distribution of the Chlorides is right skewed with a very high peak at 0.5.

```{r echo=FALSE, message=FALSE, warning=FALSE}
grid.arrange(ggplot(aes(x="pH", y= pH), data = white_wine)+geom_boxplot(),
qplot(x= pH, data= white_wine),ncol=2)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(white_wine$pH)
```

> The distribution of pH is almost symmetrical distribution.

```{r echo=FALSE, message=FALSE, warning=FALSE}
grid.arrange(ggplot(aes(x="density", y= density), data = white_wine)+geom_boxplot(),
qplot(x= density, data= white_wine),ncol=2)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(white_wine$density)
```

> The density distribution is almost symmetrical with some very extreme outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE}
grid.arrange(ggplot(aes(x="free.sulfur.dioxide", y= free.sulfur.dioxide), data = white_wine)+geom_boxplot(),
qplot(x= free.sulfur.dioxide, data= white_wine),ncol=2)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(white_wine$free.sulfur.dioxide)
```

> The distribution of the Free Sulfur dioxide is right skewed with a very high outlier at 289. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
grid.arrange(ggplot(aes(x="total.sulfur.dioxide", y= total.sulfur.dioxide), data = white_wine)+geom_boxplot(),
qplot(x= total.sulfur.dioxide, data= white_wine),ncol=2)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(white_wine$total.sulfur.dioxide)
```
> Looking at "total.sulfur.dioxide" boxplot, we can see that the distribution of the Total Sulfur Dioxide is nearly symmetrical with some very high outliers.


# Univariate Analysis

### The dataset's structure:

The data set original shape was 4898 rows by 12 columns. All variables are numerical except for the only categorical variable **quality** which is the output of the other variables. After that, we added a new categorical variable **quality.buckets**. So, the number of columns changed to 13. After adding the variable, we have noticed that the vast majority of the wines are rated as a good quality.

###The main feature:

The main feature would be the "quality". So, we will see how other variables affect it. 

### Other features that will support the investigation:
I think the "alcohol" and the "residual sugar"  will have a positive effect on the quality. Also, "alcohol"  will have a negative effect on the density while the "residual sugar" will have a positive effect. I would assume that there will be a negative correlation between the "pH" and the "acidity" because higher "acidity" will mean lower "pH". 


### Added variables:
I have added the **quality.buckets** variable to group the **quality** variable into three main classifications: low, good and high. This will help us to have a better look at relation between the quality and the other attributes, because some qualities have very few number of entries.
 

### distribution of the data:
While there was some data with normal distributions like "PH", There were some variables with very huge outliers. Some of the outliers are affected by each other like the outlier in the "density" and the "residual sugar" or the "free sulfur dioxide"" and the "total sulfur dioxide". Also, the "sulfates" and the "fixed acidity" have right-skewed distribution.


# Bivariate Plots Section

I will start by creating a correlation graph to determine which variables have strong correlation and which don't. Then I will graph those with strong correlation and calculate  their correlation coefficient. Also, in order to get rid of outliers, I have put some limits for some variables.


```{r echo=FALSE, message=FALSE, warning=FALSE}
correlation.matrix<- cor(white_wine[,1:12])# creating correlation matrix and excluding the quality.buckets variable
corrplot(correlation.matrix, method="pie", type= "lower",diag= FALSE)
```

> As we can see, there are some very strong correlations, but the vast majority is very weak correlation.So, we will focus on the strong and moderate correlations.


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= quality, y= alcohol, group= quality ), data= white_wine)+geom_boxplot()
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
cor(white_wine$alcohol,white_wine$quality)
```

> symmetrical speaking, we can see as the Alcohol percentage increases, we have a higher wine quality. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= quality.buckets, y= density), data= subset(white_wine, density<1.01))+
  geom_boxplot()
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
cor(white_wine$density, white_wine$quality)
```

> Wine in the high quality category are less dense than other categories, but that maybe due to high alcohol content in high quality wines, which is causing lower density.  

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= alcohol, y= density), data= subset(white_wine, density<1.01))+
  geom_point()
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
cor(white_wine$alcohol,white_wine$density)
```

> The correlation between the Density  and Alcohol is strong positive correlation 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= residual.sugar, y= density), data =subset(white_wine, density<1.01) )+
  geom_point()
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
cor(white_wine$residual.sugar,white_wine$density)
```

> There is a strong positive correlation between Density and Residual Sugar. Also, we notice that a lot of the points are clustered approximately  between 1 and 2, which can be explained by the high peak at the Residual Sugar at the Univariate Plots Section. 


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= total.sulfur.dioxide, y= density),
       data =subset(white_wine, density<1.01 & total.sulfur.dioxide<300))+ geom_point()
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
cor(white_wine$total.sulfur.dioxide,white_wine$density)
```

> The Total Sulfur Dioxide and the density have moderate correlation of value (0.53).

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= free.sulfur.dioxide, y= total.sulfur.dioxide),
       data =subset(white_wine, free.sulfur.dioxide<125 & total.sulfur.dioxide<300))+ geom_point()
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
cor(white_wine$free.sulfur.dioxide,white_wine$total.sulfur.dioxide)
```

> This correlation exists because Total Sulfur Dioxide consist of the Free and bound SO2.


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= pH, y= fixed.acidity), data =white_wine )+ geom_point()
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
cor(white_wine$pH, white_wine$fixed.acidity)
```

> The correlation between the pH and Fixed Acidity is weak to moderate.

# Bivariate Analysis

looking at the correlation graph, we see that the **alcohol** has the strongest correlation with the **quality** variable, but still with a value of (0.44). The correlation is still considered to be a weak to moderate correlation. Also, the density has a negative weak correlation with value of (-0.31). However, if we look at the **density** vs **residual.sugar** graph, we can see that they have a strong positive correlation equal to (0.84).Also, it has a strong negative correlation of value (-0.78) with **alcohol**. In addition, the **density** have a moderate correlation of value (0.53) with **total sulfur dioxide**. Furthermore, we have a correlation of (-0.43) between **pH** and the **fixed acidity**. 
These correlations can be explained by looking at the relations between  physical and chemical properties.



# Multivariate Plots Section


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= residual.sugar, y= density ), data =subset(white_wine,density < 1))+
  geom_point(aes(color= quality.buckets))
```

> For high quality, wine we can see that they are with higher alcohol percentage. Also, high  proportion of  good quality wine has less alcohol percentage. Meanwhile, the low quality wine is not affected by alcohol percentage. It maybe not observable due to their small number. 

```{r  echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= alcohol, y= residual.sugar ), data =subset(white_wine,density < 1))+
  geom_point(aes(color= density))
```

> This is an interesting graph because it illustrates the strong correlation between the density with alcohol and residual sugar.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= total.sulfur.dioxide, y=alcohol  ),data= white_wine)+
  geom_point(aes(color= quality.buckets))
```

> Again, the high quality wines seem to be grouped at high percentage of alcohol. Furthermore, It seemes that low quality wine have less content of total sulfur dioxide.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= residual.sugar, y= alcohol), data = subset(white_wine ,residual.sugar<20))+
  geom_point(aes(color= quality.buckets))+facet_wrap(~quality.buckets)
```

>looking at the graph, we can see that there is a high number of wines that have residual sugar between 1 and 2 despite  the quality of the wine. Also, this graph maybe misleading by giving us an impression that the high quality wines are correlated to low residual sugar content if you look at the high quality group without looking at the graph as a whole.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=total.sulfur.dioxide , y= residual.sugar),
       data= subset(white_wine,density < 1&total.sulfur.dioxide <200))+
  geom_point(aes(color= density),alpha= 0.5)
```

>  We can see that as the total sulfur dioxide and residual sugar increase, the density also increases.


# Multivariate Analysis
1. High quality wines are mostly grouped at high alcohol percentage.<br/>
2. As the content of residual sugar increases, the density increases, while the opposite is true for the alcohol.<br/>
3. high qulity wines tend to have lower residual content.<br/>
4. density has a strong positive correlation with the "total sulfur dioxide" and the content of "residual sugar".<br/>



------

# Final Plots and Summary


### Plot One
```{r  echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= quality, y= alcohol, group= quality ), data= white_wine)+geom_boxplot()+
  xlab("Quality Score")+
  ylab("Alcohol Percentage")+
  ggtitle("Alcohol boxplots by Quality")
```

### Description One
Alcohol percentage have the stongest correlation with the quality scores among other variables, and in general we can see that as the alcohol percentage increases, the quality score increases. In this plot, we can see how high quality scores have high alcohol percentage. 

### Plot Two
```{r  echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= alcohol, y= density ), data =subset(white_wine,density < 1))+
  geom_point(aes(color= quality.buckets),alpha= 0.5)+ylab("density (g / cm^3)")+
  xlab("Alcohol Percentage")+
  ggtitle("Alcohol  VS Density by Quality categories")

```

### Description Two
In this plot, we see how the high quality wine are despite  with high alcohol percentage, while the good quality wines have less alcohol percentage and higher density. We can also see how strong is the negative correlation between the density and alcohol.  

### Plot Three
```{r  echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= alcohol, y= residual.sugar ), data =subset(white_wine,density < 1))+
  geom_point(aes(color= density))+
  xlab("Alcohol Percentage")+
  ylab("Residual Sugar (g / dm^3)")+
  ggtitle("Residual Sugar VS Alcohol By Density")
```

### Description Three
The density has a very strong (the strongest in the dataset) correlation with alcohol and residual sugar. We can see how those strong correlations reflected on the graph and how the density gradually decreases as the alcohol percentage increases, while the residual sugar content decreases.

------

# Reflection


To sum up, we saw how alcohol is the most related variable to the quality score. Furthermore, how the variables correlate to each other according to their physicochemical properties. There are many factors that are described and could attribute to the quality. It was hard to guess the relationships between these factors and between each factor with the quality. Regarding the quality, there were limited number of some scores, and this can obscure some  potential correlations.In the future, it would be helpful to have more variables like the country of origin and the age of the wine to see if ageing the wine will affect the quality. The study was done on one brand only. Comparing the variables among different brands might give more realistic results.


